Dewlap color variation in \textit{Anolis sagrei} is maintained between habitats within islands of the West Indies

### 2020-05-25 ms/sections/discussion.tex
nearly there

\paragraph{Dewlap coloration differs between habitat-types} We found that male dewlap coloration in \textit{A. sagrei} significantly varied between habitat-types (beach scrub bush, primary coppice forest and mangrove forest) on five islands of the West Indies: Abaco, Bimini, Cayman Brac, Little Cayman and Long Island. However, the habitat-specific variation in dewlaps was not consistent between these islands. Although those results are consistent with adaptation at a very local scale, other evolutionary drivers could be at work, including phenotypic plasticity, random drift, or historical contingency including multiple colonization events. We reject this last explanation because all of the island populations in this study are strictly monophyletic, implying a single colonization event per island (van de Schoot, unpublished thesis; \citealt{Driessens2017}). While random drift cannot be completely ruled out, we see little evidence for a role of phenotypic isolation-by-distance (spatial autocorrelation) in explaining the differences we report. While we found a significant signal of isolation-by-distance on Eleuthera, we did not detect differences in dewlap coloration between habitats on this island.

@@

\paragraph{A role of phenotypic plasticity is unlikely} Phenotypic plasticity could explain differences in coloration between habitat populations because of diet differences. The yellow, orange and red coloration in anoline dewlaps are produced by pterins and carotenoids \citep{Ortiz1962, Ortiz1962a, Ortiz1963, Ortiz1966, Macedonia2000, Steffen2007, Steffen2009}. Animals lack the ability to synthesize carotenoids, and those must therefore be found in the diet, while pterins are synthesized from nucleotides \citep{Goodwin1984, Hill2002, Hill2006}. Experimental manipulation of dietary carotenoid content showed no effect on dewlap coloration in \textit{A. sagrei} \citep{Steffen2010} nor in \textit{A. distichus} \citep{Ng2013}, another species with an orange-based dewlap. Plasticity due to differences in development (e.g. egg rearing conditions) is unlikely because dewlap coloration develops at sexual maturity \citep{Ng2013}. \citet{Cox2017} further found a high degree of heritability of dewlap coloration in \textit{A. sagrei}. These studies suggest that dewlap coloration is not a plastic trait, although transgenerational plastic effects cannot be completely ruled out by these one or two-generation common garden experiments \citep{Tariel2020}. That leaves an adaptive explanation, where dewlap color could be under differential natural and/or sexual selection in these different habitats.

\paragraph{Divergence is maintained in the face of gene flow} The small spatial scale (the median distance between island sample sites was 11.2km + mention the minimum distance at which we found differences) and the lack of geographic barriers within islands, together with the high mobility \textit{A. sagrei} individuals \citep{Kamath2018}, imply ample opportunity for extensive gene flow between the populations in this study. In agreement with this, populations from different habitats were found to not be monophyletic based on <insert marker name here> (van de Schoot et al. unpublished thesis). While gene flow is expected to erase any differences acquired by drift without selection maintaining such polymorphism \citep{Dieckmann1999}, these divergent patterns remain, which may support an adaptive explanation. Maintenance of dewlap color divergence despite gene flow has been found in \textit{A. distichus} in Hispaniola \citep{Ng2012, Ng2016} and proposed as a mechanism of reproductive isolation in the early stages of speciation \citep{Ng2011, Lambert2013, Ng2017}. \citet{Stapley2011} found that dewlap color polymorphism was maintained in the absence of genetic structure between populations of \textit{A. apletophallus} from Panama. \citet{Thorpe2002a} found that divergence in dewlap coloration matched habitat-type better than mitochondrial lineage in \textit{A. roquet} on Martinique, and a convergent pattern was found in \textit{A. trinitatis} on the featureless island of St Vincent \citep{Thorpe2002b}. Divergence in body coloration, but not dewlap coloration, was also reported in \textit{A. conspersus} on another small island, Grand Cayman \citep{Macedonia2001}. Those results suggest a high adaptive potential of the anoline dewlap.

@@

\paragraph{Sexual selection could be at play} Substantial levels of promiscuity in \textit{A. sagrei} suggest ample opportunity for female mate choice and sexual selection \citep{Kamath2018}. A number of studies suggest that characteristics of the dewlap are indicators of male quality and may therefore act as a cue in mate choice, according to the "good genes" model of sexual selection \citep{Andersson1994}. \citet{Cook2013} found lower orange reflectance in dewlaps with heavily parasitized \textit{A. brevirostris}, suggesting a trade-off in carotenoid use between the immune response and pigment deposition. \citet{Steffen2014} found that lower UV and orange-red reflectance predict contest-winning success between males. \citet{Driessens2015} further found that more yellow and red dewlaps (relative to UV) predict better body condition, and that higher yellow and UV reflectance at the margin of the dewlap predict higher hematocrit (the concentration of red blood cells), indicating a better health. Other aspects of the dewlap may be important for sexual selection, for example, dewlap size correlates with bite force \citep{Vanhooydonck2005} and sexual size dimorphism (a proxy for sexual selection; \citealt{Vanhooydonck2009}). Even though display frequency increases in the presence of females \citep{Driessens2014}, there seems to be no link between dewlap display frequency and mating success \citep{Tokarz2002, Tokarz2005} or individual quality \citep{Driessens2015}. Our report of inconsistent divergence between islands, however, is at odds with the good-genes model of sexual selection, which would predict the evolution of dewlap coloration in the same direction across the archipelago. This is consistent with \citet{Baeckens2018}, who found no link between the average island-dewlap coloration and sexual size dimorphism in \textit{A. sagrei} across the West Indies. If sexual selection is at play in our system, it more likely involves divergent female preferences that are not linked to male quality but rather to components of the environment, and may well be arbitrary (Fisherian sexual selection; \citealt{Andersson1994}).

\paragraph{Dewlap coloration could be locally adapted} Presently, we do not know the adaptive drivers of dewlap  color divergence observed in this study. Some degree of parallel evolution is usually a good indicator for an adaptive process \citep{Losos2011}, and convergent patterns of dewlap color evolution in similar environments across islands and species have been documented \citep{Thorpe2002a, Thorpe2002b}. Here, we found evidence for a few consistencies in within-island divergence across the nine islands studied (show some archipelago-wide results?); dewlap brightness was lower in the primary coppice habitat on three islands, and the primary coppice habitat was intermediate in coloration between mangrove and beach scrub on two islands. Those patterns could reflect adaptation to components of the habitat \citep{Endler1988}. However, it is not clear what those components of the habitat might be. Previous studies have found that dewlap coloration maximizes detectability given the light conditions in the local habitat, primarily through UV contrast -- with UV-brighter dewlaps in UV-dark, mesic habitats and UV-darker dewlaps in UV-bright, xeric habitats -- in \textit{A. cristatellus} and \textit{A. cooki} on Puerto Rico \citep{Leal2002, Leal2004}. We found no such pattern in \textit{A. sagrei}, where instead, we found the darkest dewlaps in the darkest, mesic habitat -- primary coppice forest -- on three islands, and dewlaps often differed the most between beach scrub and mangrove forest, two xeric habitats with similar, high irradiance levels \citep{Howard1950, Schoener1968}. The inconsistent and idiosyncratic patterns we observed suggest that dewlap color variation between habitats cannot be predicted by habitat identity alone. Studies of Jamaican and Hispaniolan anoles similarly found between-habitat differences in dewlap coloration but no evidence for higher detectability \citep{Fleishman2009, Ng2012}. Habitats on different islands may also differ in other aspects than light conditions such as densities of predators or other anole species, which have been shown to affect among-island dewlap diversity \citep{Vanhooydonck2009, Baeckens2018}. In particular, \citet{Baeckens2018} recently showed that dewlaps with spotted patterns occurred more often in \textit{A. sagrei} on islands with more coexisting species of anoles.

>>>>

\paragraph{Dewlap coloration differs between habitat-types} We found that male dewlap coloration in \textit{A. sagrei} significantly varied between habitat-types (beach scrub bush, primary coppice forest and mangrove forest) on five islands of the West Indies: Abaco, Bimini, Cayman Brac, Little Cayman and Long Island. However, the habitat-specific variation in dewlaps was not consistent between these islands. Although those results are consistent with adaptation at a very local scale, other evolutionary drivers could be at work, including phenotypic plasticity, random drift, or historical contingency, including multiple colonization events. We reject this last explanation because all of the island populations in this study are strictly monophyletic, implying a single colonization event per island (van de Schoot, unpublished thesis; \citealt{Driessens2017}). While random drift cannot be completely ruled out, we see little evidence for a role of phenotypic isolation-by-distance (spatial autocorrelation) in explaining the differences we report. We did detect a significant signal of isolation-by-distance on Eleuthera, but we did not find differences in dewlap coloration between habitats on this island.

\paragraph{Divergence is maintained in the face of gene flow} We detected differences between habitats at relatively small spatial scales, most of the time between sites 5-10km apart, and sometimes a few hundred meters away. This, together with the lack of geographic barriers within islands and the high mobility of \textit{A. sagrei} individuals \citep{Kamath2018}, imply ample opportunity for extensive gene flow between our study populations. In agreement with this, populations from different habitats were found to not be monophyletic based on <insert marker name here> (van de Schoot et al. unpublished thesis). The maintenance of phenotypic differences at such a small spatial scales, in the face of gene flow, may support an adaptive explanation \citep{Richardson2014}.\\

The persistence of dewlap color divergence despite gene flow has been documented in multiple species of anoles, sometimes in relation to environmental conditions. \citet{Ng2012} and \citet{Ng2016} found divergent dewlap coloration in the face of gene flow between subspecies of \textit{A. distichus} across Hispaniola, and proposed this as a mechanism of reproductive isolation in the early stages of speciation \citep{Ng2011, Lambert2013, Ng2017}. \citet{Stapley2011} found that dewlap color polymorphism was maintained in the absence of genetic structure between populations of \textit{A. apletophallus} from Panama. \citet{Thorpe2002a} found that divergence in dewlap coloration matched habitat-type better than mitochondrial lineage in \textit{A. roquet} on Martinique, and a convergent pattern was found in \textit{A. trinitatis} on the featureless island of St Vincent \citep{Thorpe2002b}. Finally, divergence in body coloration, but not dewlap coloration, was also reported in \textit{A. conspersus} on another small island, Grand Cayman \citep{Macedonia2001}. Altogether, these results are consistent with a generally high adaptive potential of dewlap coloration in anoles over small distances, which could apply to $A. sagrei$.

\paragraph{Dewlap coloration could be locally adapted} Presently, we do not know the adaptive drivers of dewlap color divergence observed in this study. Some degree of parallel evolution is usually a good indicator for an adaptive process \citep{Losos2011}, and convergent patterns of dewlap color evolution in similar environments across islands and species have been documented in other species \citep{Thorpe2002a, Thorpe2002b}. Here, however, we found little consistency in between-habitat divergence across the nine islands studied; except maybe for dewlap brightness being lower in the primary coppice habitat on three islands, and the primary coppice habitat being intermediate in coloration between mangrove and beach scrub on two islands. Those patterns could reflect adaptation to components of the habitat \citep{Endler1988}, but it is not clear what those components of the habitat might be.\\

Previous studies have suggested that dewlap coloration may have evolved to be maximally detectable under local light conditions imposed by the environment, primarily through UV contrast (i.e. UV-brighter dewlaps in UV-dark, mesic habitats and UV-darker dewlaps in UV-bright, xeric habitats), in \textit{A. cristatellus} and \textit{A. cooki} from Puerto Rico \citep{Leal2002, Leal2004}. Although UV reflectance was commonly involved in beetween-habitat divergence, we found no such patterns in \textit{A. sagrei}, where instead, we found the darkest dewlaps in the darkest, mesic habitat -- primary coppice forest -- on three islands, and dewlaps often differed the most between beach scrub and mangrove forest, two xeric habitats with similar, high irradiance levels \citep{Howard1950, Schoener1968}. The inconsistent and idiosyncratic patterns we observed suggest that dewlap color variation between habitats cannot be predicted by habitat identity alone. Studies of Jamaican and Hispaniolan anoles similarly found between-habitat differences in dewlap coloration but no evidence for higher detectability \citep{Fleishman2009, Ng2012}. Habitats on different islands may also differ in other aspects than light conditions such as densities of predators or other anole species, which have been shown to affect among-island dewlap diversity \citep{Vanhooydonck2009, Baeckens2018}. In particular, \citet{Baeckens2018} recently showed that dewlaps with spotted patterns occurred more often in \textit{A. sagrei} on islands with more coexisting species of anoles. Therefore, dewlap coloration in $A. sagrei$ may be under the influence of a mosaic of interacting local environmental factors, which need not be the same across islands.

@@

\paragraph{A role of phenotypic plasticity is unlikely} Phenotypic plasticity could explain differences in coloration between habitat populations because of diet differences. The yellow, orange and red coloration in anoline dewlaps are produced by pterins and carotenoids \citep{Ortiz1962, Ortiz1962a, Ortiz1963, Ortiz1966, Macedonia2000, Steffen2007, Steffen2009}. Animals lack the ability to synthesize carotenoids, and those must therefore be found in the diet, while pterins are synthesized from nucleotides \citep{Goodwin1984, Hill2002, Hill2006}. However, experimental manipulation of dietary carotenoid content showed no effect on dewlap coloration in \textit{A. sagrei} \citep{Steffen2010} nor in \textit{A. distichus} \citep{Ng2013}, which also has an orange-based dewlap. This makes a plastic response to differences in diet across habitats unlikely. Developmental plasticity (e.g. due to differences egg rearing conditions) is also unlikely because dewlap coloration develops at sexual maturity in anoles \citep{Ng2013}. The differences we observed could therefore be heritable. This hypothesis is further supported by \citet{Cox2017}, who found a high degree of heritability of dewlap coloration in \textit{A. sagrei}. Although most studies used one or two-generation common garden experiments and thus could not rule out transgenerational plastic effects \citep{Tariel2020}, dewlap coloration generally seems to not be a plastic trait. This further reinforces an adaptive explanation, where dewlap color could be under differential natural and/or sexual selection in these different habitats.

@@

\paragraph{Sexual selection could be at play} Substantial levels of promiscuity in \textit{A. sagrei} suggest ample opportunity for female mate choice and sexual selection \citep{Kamath2018}. A number of studies have suggested that characteristics of the dewlap are indicators of male quality and may therefore act as a cue in mate choice, according to the "good genes" model of sexual selection \citep{Andersson1994}. \citet{Cook2013} found lower orange reflectance in dewlaps with heavily parasitized \textit{A. brevirostris}, suggesting a trade-off in carotenoid use between the immune response and pigment deposition. \citet{Steffen2014} found that lower UV and orange-red reflectance predict contest-winning success between males. \citet{Driessens2015} further found that more yellow and red dewlaps (relative to UV) predict better body condition, and that higher yellow and UV reflectance at the margin of the dewlap predict higher hematocrit (the concentration of red blood cells), indicating a better health.\\

Other aspects of the dewlap may be important for sexual selection. For example, dewlap size in $A. sagrei$ was found to correlate with bite force \citep{Vanhooydonck2005} and sexual size dimorphism (a proxy for sexual selection; \citealt{Vanhooydonck2009}). Even though display frequency increases in the presence of females \citep{Driessens2014}, no link was found between dewlap display frequency and mating success \citep{Tokarz2002, Tokarz2005} or individual quality \citep{Driessens2015}. Our report of inconsistent divergence between islands, however, is at odds with the good-genes model of sexual selection, which would predict the evolution of dewlap coloration in the same direction across the archipelago. This is in agreement with \citet{Baeckens2018}, who found no link between the average island-dewlap coloration and sexual size dimorphism in \textit{A. sagrei} across the West Indies. If sexual selection is at play in our system, it more likely involves divergent female preferences that are not linked to male quality but rather to components of the environment, or may well be arbitrary (Fisherian sexual selection; \citealt{Andersson1994}).

### 2020-05-25 ms/sections/results.tex
nearly there

The small archipelago-wide effect of habitat-type was detected for PC1, PC2 and PC3 (mixed-effect ANOVA with island as a random effect, Table \ref{suptab:anova_pooled}, backed up by KW), but this effect was too small for post hoc tests to find which habitats differed. Archipelago-wide differences in dewlap coloration between habitats were also detected by SVMs trained on pooled data regardless of island identity, both for PCA data and reflectance scores (Fig. \ref{supfig:classif_svm_pca_pooled}, \ref{supfig:classif_svm_refl_pooled}). This pattern seemed to be driven by mangrove lizards being correctly reassigned more often than predicted by chance. Sensitivity analyses on these machines suggest a relatively small role of long wavelengths (red reflectance) in driving this pattern (Fig. \ref{supfig:importance_svm_refl_pooled}), but did not reveal strong differences between the PCs in relative importance (Fig. \ref{supfig:importance_svm_pca_pooled}).  Archipelago-wide differences were not detected by LDA classifiers at all (Fig. \ref{supfig:classif_lda_pca_pooled}, \ref{supfig:classif_lda_refl_pooled}). So the difference could be in something else than means.\\

Within islands, SVM classifiers correctly assigned individuals to their habitat of origin based solely upon dewlap coloration on five islands: Abaco, Bimini, Cayman Brac, Little Cayman, and Long island (Fig. \ref{fig:classif_svm_pca}). An LDA approach yielded similar success rates (Fig. \ref{supfig:classif_lda_pca}), suggesting robust differences between these populations. Of the five islands, Little Cayman was the best discriminated with a mean SVM generalization success of 73.4\% (Table \ref{suptab:classif_svm_pca}). The results of the classification analyses on PCA data were very similar to results from SVMs and LDAs trained on reflectance values at 50nm-spaced wavelengths from 300 to 700nm (Fig. \ref{supfig:classif_svm_refl} and \ref{supfig:classif_lda_refl}).\\

Differentiation in dewlap coloration occurred in multiple dimensions of color space. Moreover, the differences in dewlaps between habitats were not always consistent between islands, thus, we will discuss the habitat-specific variation in dewlap coloration for each island where significant differences were detected in turn (Fig. \ref{fig:boxplots}, Table \ref{tab:anova}, Table KW). Figure \ref{fig:boxplots}A provides a key to map principal component scores to the underlying wavelengths.\\

@@

Sensitivity analyses on classifiers suggested an overall higher relative importance for PC2 and PC3 in determining between-group differences on Abaco, both in SVM and LDA classifiers (Fig. \ref{supfig:importance_svm_pca}, \ref{supfig:importance_lda_pca}), consistent with our ANOVA results (Fig. \ref{fig:boxplots}B). There was no strong signal of differences in relative importance among principal components on the other islands. Sensitivity analyses of SVMs trained on reflectance scores rather than principal components revealed, however, a consistently higher importance of ultraviolet reflectance in between-group differences on all islands (Fig. \ref{fig:importance_svm_refl}). This pattern was not recovered for LDAs trained on reflectance scores (Fig. \ref{fig:importance_lda_refl}).\\

@@

We did not find significant spatial autocorrelation between the sampling sites on the islands where we detected a significant habitat effect. We did, however, detect a significant positive signal of autocorrelation on Eleuthera ($P = 0.02$, Table \ref{suptab:autocor}), suggesting possible color differentiation through isolation-by-distance on this island.\\

@@

Patterns of differentiation were inconsistent across the five most significant islands. Contrasts in principal components between habitats, calculated on pooled data from the whole archipelago, were not similar, for any component, among islands (contrast figure, MANOVA, P > 0.05). No pattern of variation was shared by all five significant islands, along any dimension. Some patterns did seem more common however, such as darker dewlaps among coppice lizards (Cayman Brac, Little Cayman, and Long Island, Fig. \ref{fig:boxplots}) or the intermediate position of coppice lizards in chromatic color space (Cayman Brac and Long Island). In other cases, patterns of differentiation were reversed from one island to another, with more ultraviolet reflecting dewlaps in mangroves than in coastal habitat on Abaco and Cayman Brac, but the opposite on Little Cayman and Long Island. Overall, it seemed that patterns of heterogeneity of variance were often driven by higher variances in coloration within beach scrub lizards (Fig. \ref{fig:boxplots}, Table \ref{tab:anova}). Yet other patterns were idiosyncratic, such as the combination of higher red and ultraviolet reflectance in coastal lizards on Little Cayman, where the rule seemed to be a negative correlation between ultraviolet and red reflectance across every other island.\\

>>>>

The small archipelago-wide effect of habitat-type was detected for PC1, PC2 and PC3 (mixed-effect ANOVA with island as a random effect, Table \ref{suptab:anova-pooled}), but this effect was too small for post hoc tests to find which habitats differed. Archipelago-wide differences in dewlap coloration between habitats were also detected by SVMs trained on pooled data regardless of island identity, both for PCA data and reflectance scores (Fig. \ref{supfig:classif-svm-pca-pooled}, \ref{supfig:classif-svm-refl-pooled}). This pattern seemed to be driven by mangrove lizards being correctly reassigned more often than predicted by chance. Sensitivity analyses on these machines suggest a relatively small role of long wavelengths (red reflectance) in driving this pattern (Fig. \ref{supfig:importance-svm-refl-pooled}), but did not reveal strong differences between the PCs in relative importance (Fig. \ref{supfig:importance-svm-pca-pooled}).  Archipelago-wide differences were not detected by LDA classifiers at all (Fig. \ref{supfig:classif-lda-pca-pooled}, \ref{supfig:classif-lda-refl-pooled}). So the difference could be in something else than means.\\

Within islands, SVM classifiers correctly assigned individuals to their habitat of origin based solely upon dewlap coloration on five islands: Abaco, Bimini, Cayman Brac, Little Cayman, and Long island (Fig. \ref{fig:classif-svm-pca}). An LDA approach yielded similar success rates (Fig. \ref{supfig:classif-lda-pca}), suggesting robust differences between these populations. Of the five islands, Little Cayman was the best discriminated with a mean SVM generalization success of 73.4\% (Table \ref{suptab:classif-svm-pca}). The results of the classification analyses on PCA data were very similar to results from SVMs and LDAs trained on reflectance values at 50nm-spaced wavelengths from 300 to 700nm (Fig. \ref{supfig:classif-svm-refl} and \ref{supfig:classif-lda-refl}).\\

Differentiation in dewlap coloration occurred in multiple dimensions of color space. Moreover, the differences in dewlaps between habitats were not always consistent between islands, thus, we will discuss the habitat-specific variation in dewlap coloration for each island where significant differences were detected in turn (Fig. \ref{fig:anova}, Tables \ref{tab:anova}, \ref{suptab:kruskal}). Figure \ref{fig:anova}A provides a key to map principal component scores to the underlying wavelengths.\\

@@

Sensitivity analyses on classifiers suggested an overall higher relative importance for PC2 and PC3 in determining between-group differences on Abaco, both in SVM and LDA classifiers (Fig. \ref{supfig:importance-svm-pca}, \ref{supfig:importance-lda-pca}), consistent with our ANOVA results (Fig. \ref{fig:anova}B). There was no strong signal of differences in relative importance among principal components on the other islands. Sensitivity analyses of SVMs trained on reflectance scores rather than principal components revealed, however, a consistently higher importance of ultraviolet reflectance in between-group differences on all islands (Fig. \ref{supfig:importance-svm-refl}). This pattern was not recovered for LDAs trained on reflectance scores (Fig. \ref{supfig:importance-lda-refl}).\\

@@

We did not find significant spatial autocorrelation between the sampling sites on the islands where we detected a significant habitat effect. We did, however, detect a significant positive signal of autocorrelation on Eleuthera (Table \ref{suptab:autocor}), suggesting possible color differentiation through isolation-by-distance on this island.\\

@@

Patterns of differentiation were inconsistent across the five most significant islands. Contrasts in principal components between habitats, calculated on pooled data from the whole archipelago, were not similar, for any component, among islands (contrast figure, MANOVA, P > 0.05). No pattern of variation was shared by all five significant islands, along any dimension. Some patterns did seem more common however, such as darker dewlaps among coppice lizards (Cayman Brac, Little Cayman, and Long Island, Fig. \ref{fig:anova}) or the intermediate position of coppice lizards in chromatic color space (Cayman Brac and Long Island). In other cases, patterns of differentiation were reversed from one island to another, with more ultraviolet reflecting dewlaps in mangroves than in coastal habitat on Abaco and Cayman Brac, but the opposite on Little Cayman and Long Island. Overall, it seemed that patterns of heterogeneity of variance were often driven by higher variances in coloration within beach scrub lizards (Fig. \ref{fig:anova}, Table \ref{tab:anova}). Yet other patterns were idiosyncratic, such as the combination of higher red and ultraviolet reflectance in coastal lizards on Little Cayman, where the rule seemed to be a negative correlation between ultraviolet and red reflectance across every other island.\\
